---

title       : Machine Learning with R - II
author      : Ilan Man
job         : Strategy Operations  @ Squarespace
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : mathjax       # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}

----

## Agenda 
<space>

1. Logistic Regression
2. Principle Component Analysis
3. Clustering
4. Trees

----

## Objectives 
<space>

1. Understand some popular algorithms and techniques
2. Learn how to tune parameters
3. Practice R

----

## Logistic Regression
# Motivation
<space>


----

## Logistic Regression
# Motivation
<space>

```{r}
library("MASS")
library(ggplot2)
data(menarche)
log_data <- data.frame(Y=menarche$Menarche/menarche$Total)
log_data$X <- menarche$Age

glm.out <- glm(cbind(Menarche, Total-Menarche) ~ Age,family=binomial(logit), data=menarche)
lm.out <- lm(Y~X, data=log_data)

log_data$fitted <- glm.out$fitted

data_points <- ggplot(log_data) + geom_point(aes(x=X,y=Y),color='blue',size=3)
line_points <- data_points + geom_abline(intercept = coef(lm.out)[1], slope = coef(lm.out)[2],color='green',size=1)
curve_points <- line_points + geom_line(aes(x=X,y=fitted),color='red',size=1) 
```

----

## Logistic Regression
# Notation
<space>

- Binary outcome: 1 or 0
- want function bounded between 1 and 0
  - sigmoid or logistic function: $g_(z) = \frac{1}{1+\e^{-z}}$
  
----

## Logistic Regression
# Find parameters
<space>

- Goal remains the same: Minimize MSE
- set cost function to be large when guess is wrong - either guess 1 and actual is 0 or vice verse

cost = -log(x), when predict y = 1
cost = log(1-x) when predict y = 0

curve(-log(x),from = 0, to = 100)
curve(-log(1-x),from = -100, to = 0)



- $\theta$ is $\{\beta_{0}, \beta{1}\}$, where $\beta_{0}$ is $\alpha$ from before
- $X = \{x_{0}, x_{1}\}$
- Make $x_{0}$ = 1
- $h_{\theta}(x) = \beta \times t(X)$
- $J(\theta_{0},\theta_{1}) = \frac{1}{2m}\sum_{i=1}^{m}(h_{\theta}(x_{i}) - y_{i})^2$


----

## Logistic Regression
# Motivation
<space>

- Re-arranging $Y = \frac{1}{1+\e^{-\theta \times X}}$ yields
$\ln{\frac{Y}{1 - Y)}} = \theta \times X$

curve(1/(1+exp(-x)),from = -10,to = 10)

----

## Regression example
# Gradient descent
<space>

```{r grad_ex_plot,echo=FALSE,fig.height=5,fig.width=5}
x <- seq(0,1,by=0.01)
y <- 4 + 3*x*rnorm(length(x),2,sd=0.3)
df <- data.frame(cbind(X=x,Y=y))

## take a look at the data
ggplot(data=df,aes(x=X,y=Y))+geom_point(size=2)
```

----

## Regression example
# Gradient descent
<space>

```{r grad_formula,echo=TRUE}
x <- cbind(1,x)  #Add ones to x  
theta<- c(0,0)  # initalize theta vector 
m <- nrow(x)  # Number of the observations 
grad_cost <- function(X,y,theta) return(sum(((X%*%theta)- y)^2))
```

----

## Regression example
# Gradient descent
<space>

```{r grad_function,echo=TRUE}
gradDescent<-function(X,y,theta,iterations,alpha){
  m <- length(y)
  grad <- rep(0,length(theta))
  cost.df <- data.frame(cost=0,theta=0)
  
  for (i in 1:iterations){
    h <- X%*%theta
    grad <-  (t(X)%*%(h - y))/m
    theta <- theta - alpha * grad
    cost.df <- rbind(cost.df,c(grad_cost(X,y,theta),theta))    
  }  
  
  return(list(theta,cost.df))
}
```

----

## Regression example
# Gradient descent
<space>

```{r grad_vars,comment="",fig.height=5,fig.width=5}
## initialize X, y and theta
X1<-matrix(ncol=1,nrow=nrow(df),cbind(1,df$X))
Y1<-matrix(ncol=1,nrow=nrow(df),df$Y)

init_theta<-as.matrix(c(0))
grad_cost(X1,Y1,init_theta)

iterations = 100
alpha = 0.1
results <- gradDescent(X1,Y1,init_theta,iterations,alpha)
```

----

## Regression example
# Gradient descent
<space>

```{r grad_curve,fig.height=5,fig.width=5,echo=FALSE}
theta <- results[1][[1]]
cost.df <- results[2]
cost.df <- as.data.frame(cost.df)[-1,]
ggplot(cost.df,aes(x=cost.df$theta,y=cost.df$cost)) + geom_point()+
  ylab("Cost")+xlab("Theta")+ggtitle("Cost curve")
```

----

## Regression example
# Gradient descent
<space>

```{r pred,echo=TRUE,comment=""}
grad_cost(X1,Y1,theta[[1]])
## Make some predictions
intercept <- df[df$X==0,]$Y
pred <- function (x) return(intercept+c(x)%*%theta)
new_points <- c(0.1,0.5,0.8,1.1)
new_preds <- data.frame(X=new_points,Y=sapply(new_points,pred))
```

----

## Regression example
# Gradient descent
<space>

```{r new_point,echo=TRUE,fig.height=5,fig.width=5}
ggplot(data=df,aes(x=X,y=Y))+geom_point(size=2)
ggplot(data=df,aes(x=X,y=Y))+geom_point()+geom_point(data=new_preds,aes(x=X,y=Y,color='red'),size=3)+scale_colour_discrete(guide = FALSE)
```

----

## Regression example
# Gradient descent - summary
<space>

- minimization algorithm
- approximation, non-closed form solution
- good for large number of examples
- hard to select the right $\alpha$
- traditional looping is slow - optimization algorithms are used in practice

----

## Logistic Regression
# Motivation
<space>

# example

----

## Logistic Regression
# Motivation
<space>

# Summary

----

## Principle Component Analysis
# Motivation
<space>

- used widely in modern data analysis
- not well understood
- intuition: reduce data into only relevant dimensions
- the goal of PCA is to compute the most meaningful was to re-express noisy data, revealing the hidden structure

----

## Principle Component Analysis
# Concepts
<space>

- first big assumption: linearity
- $PX=Y$
  - $X$ is original dataset, $P$ is a transformation of $X$ into $Y$
- how do we choose $P$?
  - reduce noise
  - maximize variance

----

## Principle Component Analysis
# Concepts
<space>

- covariance matrix
     - $C = X*X^{T}$

- restated goals are
  - minimize covariance and maximize variance
  - the optimizal $C$ is a diagonal matrix, off diagonals are = 0
  
----

## Principle Component Analysis
# Concepts
<space>

- summary of assumptions
  - linearity (non-linear is a kernel PCA)
  - largest variance indicates most signal, low variance = noise
  - orthogonal components - makes the linear algebra easier
  - assumes data is normally distributed, otherwise PCA might not diagonalize matrix
    - can use ICA
    - but most data is normal and PCA is robust to slight deviance from normality

----

## Principle Component Analysis
# Eigenwhat?
<space>

- $Ax = \lambdax$
  - $\lambda$ is an eigenvalue of $A$ and $x$ is an eigenvector of $A$
- $Ax - \lambdaIx = 0$
- $(A - \lambdaI)x = 0$
- $\det(A - \lambdaI)$ = 0


----

## Principle Component Analysis
# Motivation
<space>

$\[A=\left[{\begin{array}{cc}5 & 2 \\2 & 5\\\end{array}\right ]\]$

A = matrix(c(5,2,2,5),nrow=2)
I = diag(nrow(A))
|A - L*I| = 0
det(c(5-l,2,2,5-l))
(5-l)*(5-l) - 4 = 0
25 - 10l + l^2 - 4 = 0
l^2 - 10l + 21 = 0
roots <- Re(polyroot(c(21,-10,1)))
```

----

## Principle Component Analysis
# Motivation
<space>

# when lambda = -3
Ax = 3x
5x1 + 2x2 = 3x1
2x1 + 5x2 = 3x2
x1=-x2
# one eigenvector = [1 -1]. any scalar multiple of this counts.

----

## Principle Component Analysis
# Motivation
<space>

# when lambda = 7
5x1 + 2x2 = 7x1
2x2 + 5x2 = 7x2
x1 = x2
# another eigenvector = [1 1]. any scalar multiple of this counts.

----

## Principle Component Analysis
# Motivation
<space>

A%*%c(1,-1) == 3 * as.matrix(c(1,-1))
A%*%c(1,1) == 7 * as.matrix(c(1,1))
roots

----

## Principle Component Analysis
# Motivation
<space>

- check
m <- matrix(c(1,-1,1,1),ncol=2)
m <- m/sqrt(norm(m))
A == as.matrix(m%*%diag(roots)%*%t(m))
# lambda is a diagonal matrix, with 0 off diagonals

----

## Principle Component Analysis
# Motivation
<space>

PX = Y

CY = (1/(n-1))*YYt
=PX(PX)t
=PXXtPt
=PAPt
# P is a matrix with columns that are eigenvectors
# A is a diagonalized matrix of eigenvalues (by linear algebra) and symmetric
A = EDEt

----

## Principle Component Analysis
# Motivation
<space>

# each row of P should be an eigenvector of A
P=Et
# also note that Pt = P-1 (linear algebra)
A = PtDP
CY = PPtDPPt
= (1/(n-1))*D
# D is a diagonal matrix, depending on how we choose P
# therefore CY is diagonalized

----

## Principle Component Analysis
# Example
<space>

data <- read.csv('tennis_data_2013.csv')
data$Player1 <- as.character(data$Player1)
data$Player2 <- as.character(data$Player2)

tennis <- data
m <- length(data)

for (i in 10:m){
  tennis[,i] <- ifelse(is.na(data[,i]),0,data[,i])
}

str(tennis)

features <- tennis[,10:m]

head(features)
str(features)
dim(features)

----

## Principle Component Analysis
# Example
<space>

scaled_features <- as.matrix(scale(features))
Cx <- cov(scaled_features)
eigenvalues <- eigen(Cx)$values
eigenvectors <- eigen(Cx)$vectors
PC <- scaled_features %*% eigenvectors

----

## Principle Component Analysis
# Example
<space>

Cy <- cov(PC)
sum(round(diag(Cy) - eigenvalues,5))
sum(round(Cy[upper.tri(Cy)],5)) ## off diagonals are 0 since PC's are orthogonal

----

## Principle Component Analysis
# Example
<space>

var_explained <- round(eigenvalues/sum(eigenvalues) * 100, digits = 2)
cum_var_explained <- round(cumsum(eigenvalues)/sum(eigenvalues) * 100, digits = 2)

var_explained <- as.data.frame(var_explained)
names(var_explained) <- "variance_explained"
var_explained$PC <- as.numeric(rownames(var_explained))
var_explained <- cbind(var_explained,cum_var_explained)

library(ggplot2)
ggplot(var_explained) +
  geom_bar(aes(x=PC,y=variance_explained),stat='identity') +
  geom_line(aes(x=PC,y=cum_var_explained))

----

## Principle Component Analysis
# Example
<space>

eigenvalues = (pca.df$sdev)^2
eigenvectors are the loadings - linear combination of the variables
eigenvectors[,1] = pca.df$rotation[,1]
sum((eigenvectors[,1])^2)

----

## Principle Component Analysis
# Example
<space>

pca.df <- prcomp(scaled_features)
rows <- nrow(tennis)
pca.plot <- as.data.frame(pca.df$x[,1:2])
pca.plot$gender <- data$Gender
ggplot(data=pca.plot,aes(x=PC1,y=PC2,color=gender)) + geom_point()

----

## Principle Component Analysis
# Example
<space>

- how accurate is the first PC at dividing the dataset?
gen <- ifelse(pca.df$x[,1] > abs(mean(pca.df$x[,1]))*2,"F","M")
sum(diag(table(gen,as.character(data$Gender))))/rows

----

## Principle Component Analysis
# Summary
<space>

----

## Clustering
# Motivation
<space>

- used to separate data into meaningful or useful groups (or both)
     - capture natural structure of the data
     - useful starting point for further analysis
- customer segmentation
- cluster for utility
     - summarizing data for less expensive computation
     - data compression
     - nearest neighbors - distance between two cluster centers (centroids)
     
- types of clusters
     - data points that are more similar to one another than points outside of the cluster - most intuitive definition
     - prototype-based: each data point is more similar to the prototype, i.e. center, of the cluster than the prototype of other clusters. Often a centroid, i.e. mean.
     - density based clusters: where the density is highest, that is a cluster. Works well for data with noise and outliers. Clusters separated by noise.

----

## Clustering
# Kmeans
<space>

- k-means
     - prototype, partitional based
     - choose K initial centroids/clusters
     - points are assigned to the closest centroid
     - centroid is then updated based on the points in that cluster
     - update steps until no point changes or centroids remain the same



1. Select K points as initial centroids. 
2. repeat
3.     Form K clusters by assigning each point to its closest centroid.
4.     Recompute the centroid of each cluster. 
5. until Centroids do not change, or change very minimally, i.e. <1%


3. Use similarity measures such as Euclidean or cosine similarity depending on the data
4. Minimize the squared distance of each point to closest centroid, minimize the objective function
     - the centroid that minimizes the SSE of the cluster is the mean
     - Kmeans leads to local minimum, not global, since you’re optimizing based on the centroids you chose, not all possible centroids
     - 
- choose K randomly - can lead to poor centroids
     - run k-means multiple times - still doesn’t solve problems

- can reduce the total SSE by increasing the K
     - can increase the cluster with largest SSE
- can decrease K and minimize SSE
     - split up a cluster into other clusters. the centroid that is split will increase total SSE the least
- bisecting K means
     - less susceptible to initialization problems
     - split points into 2 clusters
          - take cluster with largest SSE - split that into two clusters
     - rerun bisecting K mean on resulting clusters
     - stop when you have K clusters

- K mean fails
     - if some clusters are much bigger than other clusters - it cannot distinguish between natural clusters
     - if clusters have different densities, K means cannot tell 
     - distance metric doesn’t account for non-globular clusters, i.e. if they follow a distribution
- K means will still work if user accepts sub clusters of the natural cluster
- strengths
     - simple, efficient computationally
     - not useful for non-globular, different density, different sized data
     - outlier detection and removal can help address outlier problem
- can derive K mean algorithm using gradient descent
     - can use calculus to show that the mean of the cluster is the best choice of centroid, i.e. minimizes SSE

----

## Clustering
# Kmeans
<space>

tennis_kmean <- kmeans(features, centers=5)

# K MEANS DOES A GOOD JOB IN CLUSTERING GENDERS
table(tennis$Gender,tennis_kmean$cluster)

----

## Clustering
# Kmeans animation
<space>

# animate kmean algorithm
install.packages('animation')
library(animation)

oopt = ani.options(interval = 1)
## the kmeans() example; very fast to converge!
ani_ex = rbind(matrix(rnorm(100, sd = 0.3), ncol = 2), 
          matrix(rnorm(100, sd = 0.3), 
          ncol = 2))
colnames(ani_ex) = c("x", "y")

kmeans.an = function(
  x = cbind(X1 = runif(50), X2 = runif(50)), centers = 4, hints = c('Move centers!', 'Find cluster?'),
  pch = 1:5, col = 1:5
) {
  x = as.matrix(x)
  ocluster = sample(centers, nrow(x), replace = TRUE)
  if (length(centers) == 1) centers = x[sample(nrow(x), centers), ] else
    centers = as.matrix(centers)
  numcent = nrow(centers)
  dst = matrix(nrow = nrow(x), ncol = numcent)
  j = 1
  pch = rep(pch, length = numcent)
  col = rep(col, length = numcent)
  
  for (j in 1:ani.options('nmax')) {
    dev.hold()
    plot(x, pch = pch[ocluster], col = col[ocluster], panel.first = grid())
    mtext(hints[1], 4)
    points(centers, pch = pch[1:numcent], cex = 3, lwd = 2, col = col[1:numcent])
    ani.pause()
    for (i in 1:numcent) {
      dst[, i] = sqrt(apply((t(t(x) - unlist(centers[i, ])))^2, 1, sum))
    }
    ncluster = apply(dst, 1, which.min)
    plot(x, type = 'n')
    mtext(hints[2], 4)
    grid()
    ocenters = centers
    for (i in 1:numcent) {
      xx = subset(x, ncluster == i)
      polygon(xx[chull(xx), ], density = 10, col = col[i], lty = 2)
      points(xx, pch = pch[i], col = col[i])
      centers[i, ] = apply(xx, 2, mean)
    }
    points(ocenters, cex = 3, col = col[1:numcent], pch = pch[1:numcent], lwd = 2)
    ani.pause()
    if (all(ncluster == ocluster)) break
    ocluster = ncluster
  }
  invisible(list(cluster = ncluster, centers = centers))
}

kmeans.an(ani_ex, centers = 5, hints = c("Move centers","Cluster found?"))

----

## Clustering
# DBSCAN
<space>

- density based
     - center based approach to finding density
     - count the number of points within some radius of a point, the radius is call Eps
     - if Eps is too big, there will be m points, if eps is too small, there will be 1 point
     - core point has X points within a radius of Eps, border points are within a radius of Eps of core point, and noise points are not within Eps of border or core points
     - if p is density connected to q, they are part of the same cluster, if not, then they are not; if p is not density connected to any other point, its considered noise
     
----

## Clustering
# DBSCAN
<space>

x <- c(2,2,8,5,7,6,1,4)
y <- c(10,5,4,8,5,4,2,9)
cluster <- data.frame(X=c(x,2*x,3*x),Y=c(y,-2*x,1/4*y))
plot(cluster)

----

## Clustering
# DBSCAN
<space>

install.packages('fpc')
library(fpc)
cluster_DBSCAN<-dbscan(cluster, eps=3, MinPts=2, method="hybrid")
plot(cluster_DBSCAN, cluster, main="Clustering using DBSCAN algorithm (eps=3, MinPts=3)")

----

## Clustering
# Summary
<space>

----

## Trees
# Motivation
<space>

- A root node that has no incoming edges and zero or more outgoing edges.
- Internal nodes, each of which has exactly one incoming edge and two or more outgoing edges
- Leaf or terminal nodes, each of which has exactly one incoming edge and no outgoing edges. 

- The non- terminal nodes, which include the root and other internal nodes, contain attribute test conditions to separate records that have different characteristics

- trees work best with categorical values
- descriptions are disjoint
- trees are robust to data errors
- training data is missing values

----

## Trees
# Algorithm(s)
<space>

ID3
C4.5
C5.0

----

## Trees
# Entropy calculation
<space>


- The entropy of a sample of data indicates how mixed the class values are; the minimum value of 0 indicates that the sample is completely homogenous, while 1 indicates the maximum amount of disorder.
     - draw curve for entropy
- InfoGain = Entropy (pre split) - Entropy (post split)
     - Entropy is weighted by the Entropy of each feature split
 - avoid pre-pruning because its impossible to know if the tree will miss subtle but important patterns in the data (if you prune too early)
- hard to know optimal length of tree without growing it there first

- Entropy = expected amount of information contained in a random variable -> information is synonymous with "bits" which is why is log, base 2 
     - the more a feature splits the data in obvious ways, the less informative it is for us, entropy is lower
     - the more the feature splits the data, the higher the entropy and hence information gained by splitting at that feature
     - Entropy is minimized when one of the events has a P(X)=1
     - Entropy is maximized when each event has a P(X)=1/n of happening


----

## Trees
# Entropy calculation
<space>

entropy_function <- function(p)
  {
    if (min(p) < 0 || sum(p) <= 0)
      return(NA)
    p.norm <- p[p>0]/sum(p)
    -sum(log2(p.norm)*p.norm)
  }

entropy_function(c(0.99,1))

----

## Trees
# Example
<space>

voting_data <- read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/voting-records/house-votes-84.data')
str(voting_data)
names(voting_data) <- c('party','handicapped-infants','water-project-cost-sharing',
                        'adoption-of-the-budget-resolution','physician-fee-freeze',
                        'el-salvador-aid','religious-groups-in-schools',
                        'anti-satellite-test-ban','aid-to-nicaraguan-contras',
                        'mx-missile','immigration','synfuels-corporation-cutback',
                        'education-spending','superfund-right-to-sue','crime',
                        'duty-free-exports','export-administration-act-south-africa')


prop.table(table(voting_data[,1]))
n <- nrow(voting_data)
train_ind <- sample(n,2/3*n)
voting_train <- voting_data[train_ind,]
voting_test <- voting_data[-train_ind,]

----

## Trees
# Example
<space>

install.packages("party")
library(C50)
library(party)
library(gmodels)
head(voting_train)

# plot tree using party package
tree_formula <- with(voting_train,voting_train$party ~ .)
p_tree <- ctree(tree_formula,data=voting_train)
plot(p_tree,
     inner_panel=node_inner(p_tree,pval = FALSE,id = TRUE),
     terminal_panel=node_terminal(p_tree, digits = 1, id = TRUE))

----

## Trees
# Example
<space>

# make tree using C5.0
tree_model <- C5.0(voting_train[,-1],voting_train[,1], trials=1)
tree_predict <- predict(tree_model, newdata=voting_test[,-1])
conf <- CrossTable(voting_test[,1], tree_predict, prop.chisq = FALSE,
                   prop.c = FALSE, prop.r = FALSE,
                   dnn = c("actual class", "predicted class"))

----

## Trees
# Example
<space>

# most important variables
C5imp(tree_model)

# in-sample error rate
summary(tree_model)

----

## Trees
# Example - Boosting
<space>

# boosting is rooted in the notion that by combining a number of weak performing learners, 
# you can create a team that is much stronger than any one of the learners alone.
# this is where C5.0 improves on C4.5

----

## Trees
# Example - Boosting
<space>

boosted_tree_model <- C5.0(voting_train[,-1],voting_train[,1], trials=25)
boosted_tennis_predict <- predict(boosted_tree_model,voting_test[,-1])

boosted_conf <- CrossTable(voting_test[,1], boosted_tennis_predict, prop.chisq = FALSE,
                           prop.c = FALSE, prop.r = FALSE, 
                           dnn = c("actual class", "predicted class"))

# in-sample error rate
summary(boosted_tree_model)

----

## Trees
# Example - Error Cost
<space>

# still getting too many false positives (predict republican but actually democrat)
# introduce higher cost to getting this wrong

error_cost <- matrix(c(0,1,2,0),nrow=2)
cost_model <- C5.0(voting_train[,-1],voting_train[,1], trials=1, costs = error_cost)
cost_predict <- predict(cost_model, newdata=voting_test[,-1])
conf <- CrossTable(voting_test[,1], cost_predict, prop.chisq = FALSE,
                   prop.c = FALSE, prop.r = FALSE,
                   dnn = c("actual class", "predicted class"))

----

## Trees
# Example - Error Cost
<space>

tris <- seq(1,50,by=2)
boost_acc <- NULL
for (i in tris){  
  temp <- C5.0(voting_train[,-1],voting_train[,1], trials=i, costs = error_cost)
  temp_pred <- predict(temp,voting_test[,-1])
  boost_acc <- append(boost_acc,sum(diag(table(temp_pred,voting_test[,1]))))
}

plot(boost_acc,type='l')

----

## Trees
# Pros and Cons
<space>


- trees are non-parametric, rule based classification or regression method
- simple to understand and interpret
- little data preparation
- easy to overfit (need to prune to avoid that, or have max tree depth)
- the global optimum is known as NP-complete unless greedy search is employed, which is then only locally optimum. Can mitigate this with an ensemble of trees
- difficult concepts that are not easily expressed by trees (XOR) are hard to learn
- for class imbalance, trees can be biased - should balance dataset before fitting
- trees tend to overfit, so use PCA beforehand

----

## Resources
<space>

- [Machine Learning with R](http://www.packtpub.com/machine-learning-with-r/book)
- [Machine Learning for Hackers](http://shop.oreilly.com/product/0636920018483.do)
- [Elements of Statistical Learning](http://web.stanford.edu/~hastie/local.ftp/Springer/OLD/ESLII_print4.pdf)

----